### 2020-01-29 wuhan.tex
stuff

As the novel coronavirus (2019-nCoV) continues to spread,

@@

Since December 2019, the novel coronavirus (2019-nCoV) has been
spreading in China and in other parts of the world.
As of January 28th, 2020, the World Health Organization (WHO) has
confirmed 4593 cases, including 56 confirmed cases in 14 different
countries, outside China \citep{who28report}.
Although the virus is believed to have originated
from animal reservoirs, a recent case from Viet Nam
demonstrated its ability to directly transmit between
humans \citep{who26report},
posing a greater threat for its spread.

Many researchers have been rushing to publish their 
analysis of the outbreak and, in particular, their
estimates of the basic reproductive number $\mathcal R_0$ (i.e., the 
average number of secondary cases generated 
by a primary case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}).
The basic reproductive number is of particular interest 
because it allows prediction about the final size of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.
While their efforts are valuable, their analyses rely on
assumptions that could immediately affect their estimates of $\mathcal R_0$ and
the associated uncertainties.

Here, we present a statistical framework for evaluating uncertainties 
associated with parameter estimates
across a wide range of models.
Our results indicate that most early estimates of $\mathcal R_0$ for the outbreak
are likely to be overly confident.

@@

Here, we use the gamma approximation framework \citep{park2019practical} to first characterize the
amount of uncertainty present in the underlying parameters and then assess the 
degree to which these uncertainties
affect the estimate of $\mathcal R_0$.
Assuming that generation intervals follow a gamma distribution 
with the mean $\bar G$ and the squared coefficient of variation (CV$^2$) $\kappa$, 
we have
\begin{equation}
\mathcal R_0 = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
This equation demonstrates that a generation-interval distribution
that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$)
will give a higher estimate of $\mathcal R_0$ for the same value of $r$.

@@

To capture the average ranges of uncertainties for each parameter ($r$, $\bar G$, and $\kappa$),
we use a Bayesian multilevel modeling approach by assuming that all \emph{observed} (i.e., estimated or assumed) parameter values come from the same distribution:
\begin{equation}
\begin{aligned}
r_i^n &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_r, \mathrm{shape}=\alpha_r),\\
\bar{G}_i^n &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_G, \mathrm{shape}=\alpha_G),\\
\kappa_i^n &\sim \mathrm{Exponential}(\mathrm{mean}=\mu_\kappa),\\
(\mu_r, \alpha_r, \mu_G, \alpha_G, \mu_\kappa) &\sim \mathrm{Gamma}(\mathrm{shape}=0.1, \mathrm{rate}=0.1),
\end{aligned}
\end{equation}
where $n$ is randomly selected between 1 and 100,000 at each Metropolis-Hastings step;
this approach is analogous to Bayesian methods for analyzing phylogenetic data, which can rely on drawing random samples of phylogenetic trees from a discrete set to account for phylogenetic uncertainty \citep{pagel2004bayesian,bedford2014integrating}.
We use an exponential distribution to model $\kappa_i$ since one model assumes a fixed generation interval ($\kappa = 0$).
We run 4 parallel chains that consist of 100,000 burnin steps and 100,000 sampling steps.
Posterior samples are thinned every 200 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all parameters \citep{gelman1992inference}.

@@

\fref{assumption} compares the estimated (or assumed) values of the exponential growth rate $r$, mean generation interval $\bar G$, and the squared CV $\kappa$ of the generation-interval distribution with the average values ($\mu_r$, $\mu_G$, and $\mu_\kappa$) that we calculate from our multilevel model.
We find that there is a large uncertainty associated with the underlying parameters;
many models rely on stronger assumptions that ignore these uncertainties.
Surprisingly, no studies take into account how the variability in generation intervals (represented by $\kappa$) affects their estimates of $\mathcal R_0$, even though the assumed values range from 0 to 0.5 across different models.

To evaluate how the uncertainty in each parameter affects the estimate of $\mathcal R_0$,
we replace the estimated (or assumed) values of $r$, $\bar G$, and $\kappa$ with the average values ($\mu_r$, $\mu_G$, and $\mu_\kappa$) one at a time and recalculate the basic reproductive number $\mathcal R_0$ (\fref{R0}).
We find that incorporating uncertainties one at a time increases the width of the confidence intervals in most cases (15 out 18).
We estimate narrower confidence intervals for Study 2 and Study 3 when we include uncertainties in $\kappa$ because they assume narrow generation-interval distributions ($\kappa = 0.2$ and $\kappa=0$, respectively);
when lower values of $\kappa$ are used, their estimates of $\mathcal R_0$ become less sensitive to the values of $r$ and $\bar G$.
We estimate narrower confidence intervals when we use the average mean generation time $\mu_G$ to recalculate $\mathcal R_0$ for Study 5 because the range of uncertainty in the mean generation time $\bar G$ that they consider is much wider than ours (\fref{assumption}).

We also find that accounting for uncertainties in the estimate of $r$ has the largest effect on the estimates of $\mathcal R_0$ and the width of associated confidence intervals (\fref{R0}).
For example, recalculating $\mathcal R_0$ for Study 3 by using the average $r$ values gives $\mathcal R_0 = 4.0$ (95\% CI: 2.3--10.1), which is much wider than the range they reported (2.0--3.1).
There are two explanations for this result.
First, even though the exponential growth rate $r$ and the mean generation time $\bar G$ have identical effects on $\mathcal R_0$ under the gamma approximation framework \eref{gamma},
$r$ has greater overall effect on $\mathcal R_0$ because there is more uncertainty associated with our estimate of $\mu_r$ (95\% CI: 0.12--0.28 days$^{-1}$) than with $\mu_G$ (95\% CI: 6.6--10.2 days).
Second, assuming a fixed generation time ($\kappa=0$) makes the estimate of $\mathcal R_0$ too sensitive to $r$ and $\bar G$.

@@

Finally, we recalculate $\mathcal R_0$ by accounting for all uncertainties (i.e., using posterior samples for $\mu_r$, $\mu_G$, and $\mu_\kappa$) and compare it with the reported $\mathcal R_0$ estimates.
We estimate the average $\mathcal R_0$ to be 3.1 (95\% CI: 2.1--5.7).
While our pooled estimate of $\mathcal R_0$ is similar to other reported values, our calculation gives wider confidence intervals than all of them.
This result does not imply that our assumptions are too weak;
we believe that this confidence interval accurately reflects the level of uncertainties present in the information that were available when these models were fitted.
In fact, our results show that incorporating uncertainties in all parameters can give narrower confidence intervals than doing so one at a time because nonlinear effects of parameters could cancel out and give appropriately wide confidence intervals.

@@

Estimating the basic reproductive number $\mathcal R_0$ is crucial for predicting the course of an outbreak and planning intervention strategies.
When $\mathcal R_0$ is estimated from a wide range of models that rely on different assumptions, it can difficult to assess the statistical validity of these estimates.
Here, we present a simple framework that allows us to evaluate the validity of the estimates of $\mathcal R_0$ and associated model assumptions using the gamma approximation of the generation-interval distribution \citep{park2019practical}. 
We then apply the framework to the preliminary estimates of $\mathcal R_0$ during the novel coronavirus outbreak.
Our results demonstrates the importance of accounting for uncertainties associated with the underlying generation-interval distributions, especially with the amount of dispersion in the generation intervals:
assuming a narrow generation-interval distribution can make the estimate of $\mathcal R_0$ too sensitive to the estimates of the exponential growth rate $r$.

>>>>

@@

Since December 2019, the novel coronavirus (2019-nCoV) has been
spreading in China and in other parts of the world.
As of January 28th, 2020, the World Health Organization has
confirmed 4593 cases, including 56 confirmed cases in 14 different
countries, outside China \citep{who28report}.
Although the virus is believed to have originated
from animal reservoirs, a recent case from Viet Nam
demonstrated its ability to directly transmit between
humans \citep{who26report},
posing a greater threat for its spread.

As the disease continues to spread, 
many disease modelers have been rushing to publish their 
analysis of the outbreak and, in particular, their
estimates of the basic reproductive number $\mathcal R_0$ (i.e., the 
average number of secondary cases generated 
by a primary case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}).
The basic reproductive number is of particular interest 
because it allows prediction about the final size of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.
While their efforts are valuable, it is difficult to
compare their estimates of $\mathcal R_0$ because their methods 
and assumptions, which can directly affect their estimates and
the associated uncertainties, vary widely.

Here, we present a simple, statistical framework for evaluating uncertainties 
associated with the estimates of $\mathcal R_0$
across a wide range of models and compare estimates of $\mathcal R_0$ for the outbreak.
Our results indicate that most early estimates of $\mathcal R_0$ for the outbreak
are likely to be overly confident.

@@

Here, we use the gamma approximation framework \citep{park2019practical} to characterize the
amount of uncertainty present in the underlying parameters and assess the 
degree to which these uncertainties
affect the estimate of $\mathcal R_0$.
Assuming that generation intervals follow a gamma distribution 
with the mean $\bar G$ and the squared coefficient of variation (CV$^2$) $\kappa$, 
we have
\begin{equation}
\mathcal R_0 = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
This equation demonstrates that a generation-interval distribution
that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$)
will give a higher estimate of $\mathcal R_0$ for the same value of $r$.

@@

To capture the average ranges of uncertainties in each parameter ($r$, $\bar G$, and $\kappa$),
we use a Bayesian multilevel modeling approach by assuming that all \emph{observed} (i.e., estimated or assumed) parameter values come from the same distribution:
\begin{equation}
\begin{aligned}
r_i^n &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_r, \mathrm{shape}=\alpha_r),\\
\bar{G}_i^n &\sim \mathrm{Gamma}(\mathrm{mean}=\mu_G, \mathrm{shape}=\alpha_G),\\
\kappa_i^n &\sim \mathrm{Exponential}(\mathrm{mean}=\mu_\kappa),\\
(\mu_r, \alpha_r, \mu_G, \alpha_G, \mu_\kappa) &\sim \mathrm{Gamma}(\mathrm{shape}=0.1, \mathrm{rate}=0.1),
\end{aligned}
\end{equation}
where $n$ is randomly selected between 1 and 100,000 at each Metropolis-Hastings step;
this approach is analogous to Bayesian methods for analyzing phylogenetic data, which can rely on drawing random samples of phylogenetic trees from a discrete set to account for phylogenetic uncertainty \citep{pagel2004bayesian,bedford2014integrating}.
We use an exponential distribution to model $\kappa_i$ since one model assumes a fixed generation interval ($\kappa = 0$).
We run 4 parallel chains that consist of 100,000 burnin steps and 100,000 sampling steps.
Posterior samples are thinned every 200 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all parameters \citep{gelman1992inference}.

@@

\fref{assumption} compares the estimated (or assumed) values of the exponential growth rate $r$, mean generation interval $\bar G$, and the squared CV $\kappa$ of the generation-interval distribution with our pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$) that we calculate from our multilevel model.
We find that there is a large uncertainty associated with the underlying parameters;
many models rely on stronger assumptions that ignore these uncertainties.
Surprisingly, no studies take into account how the variability in generation intervals (represented by $\kappa$) affects their estimates of $\mathcal R_0$, even though the assumed values range from 0 to 0.5 across different studies.

To evaluate how the uncertainty in each parameter affects the estimate of $\mathcal R_0$,
we replace the estimated (or assumed) values of $r$, $\bar G$, and $\kappa$ with our pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$) one at a time and recalculate the basic reproductive number $\mathcal R_0$ (\fref{R0}).
We find that incorporating uncertainties one at a time increases the width of the confidence intervals in most cases (15 out 18).
We estimate narrower confidence intervals for Study 2 and Study 3 when we use our pooled estimate of the squared CV in generation time $\mu_\kappa$ to recalculate $\mathcal R_0$ because both studies assume narrow generation-interval distributions ($\kappa = 0.2$ and $\kappa=0$, respectively);
when lower values of $\kappa$ are used, their estimates of $\mathcal R_0$ become less sensitive to the values of $r$ and $\bar G$.
We estimate narrower confidence intervals when we use our pooled estimate of the mean generation time $\mu_G$ to recalculate $\mathcal R_0$ for Study 5 because the range of uncertainty in the mean generation time $\bar G$ that they consider is much wider than ours (\fref{assumption}).

We find that accounting for uncertainties in the estimate of $r$ (i.e., replacing the estimated $r$ values with $\mu_r$) has the largest effect on the estimates of $\mathcal R_0$ and the width of associated confidence intervals (\fref{R0}).
For example, recalculating $\mathcal R_0$ for Study 3 by using the average $r$ values gives $\mathcal R_0 = 4.0$ (95\% CI: 2.3--10.1), which is much wider than the range they reported (2.0--3.1).
There are two explanations for this result.
First, even though the exponential growth rate $r$ and the mean generation time $\bar G$ have identical effects on $\mathcal R_0$ under the gamma approximation framework \eref{gamma},
$r$ has greater overall effect on $\mathcal R_0$ because there is more uncertainty associated with our estimate of $\mu_r$ (95\% CI: 0.12--0.28 days$^{-1}$) than with $\mu_G$ (95\% CI: 6.6--10.2 days).
Second, assuming a fixed generation time ($\kappa=0$) makes the estimate of $\mathcal R_0$ too sensitive to $r$ and $\bar G$.

@@

Finally, we incorporate all uncertainties by using posterior samples for $\mu_r$, $\mu_G$, and $\mu_\kappa$ to recalculate $\mathcal R_0$ and compare it with the reported $\mathcal R_0$ estimates.
We estimate the average $\mathcal R_0$ to be 3.1 (95\% CI: 2.1--5.7).
While our pooled estimate of $\mathcal R_0$ is similar to other reported values, our calculation gives wider confidence intervals than all of them.
This result does not imply that our assumptions are too weak;
we believe that this confidence interval accurately reflects the level of uncertainties present in the information that were available when these models were fitted.
In fact, our results show that incorporating uncertainties in all parameters can give narrower confidence intervals than doing so one at a time because nonlinear effects of parameters could cancel out and give appropriately wide confidence intervals.

@@

Estimating the basic reproductive number $\mathcal R_0$ is crucial for predicting the course of an outbreak and planning intervention strategies.
Here, we evaluated the statistical validity of the estimates of $\mathcal R_0$ for the novel coronavirus outbreak using a novel statistical framework, based on the gamma approximation of the generation-interval distribution \citep{park2019practical}.
Our results demonstrate the importance of accounting for uncertainties associated with the underlying generation-interval distributions, especially with the amount of dispersion in the generation intervals:
assuming a narrow generation-interval distribution can make the estimate of $\mathcal R_0$ too sensitive to the estimates of the exponential growth rate $r$.
